Short-term synaptic and neuronal plasticity act in synergy for deviance detection in spiking networks

### 2022-12-14 paper-1/manuscript.tex
Rev sec Results.TA (F5~6)

\begin{figure*}%
    \centering
    \includegraphics{fig4.pdf}
    \caption{}
    \label{fig4}
\end{figure*}
\textbf{Fig. 4} Responses and threshold adaptation with TA only.
\textbf{a} Target trial average post-stimulus spike histogram, showing trial time along the horizontal axis, and neurons along the vertical, sorted by the time of the first recorded spike across all trials and conditions. Left and middle column, target trials in dev sequence and msc sequence, respectively; right column, contrast between these two.
\textbf{b} Target trial average of the threshold adaptation voltage $V_{TA}$, with axes, neuron order and columns as in panel a.
\textbf{c} Target trial average of the membrane potential $V_m$, with axes, neuron order and columns as in panel a. The first 10 ms of the neurons under direct stimulation are masked out to avoid saturation of the color mask.
\textbf{d} Relationship across excitatory neurons between the contrast (dev - msc) in $V_{TA}$ at the start of target trials, and the contrast in average number of spikes fired in target trials, and associated linear regression.
\textbf{e} Mean contrast (dev - msc) in $V_{TA}$ across networks and stimuli, plotted against the Pearson correlation coefficients $\rho$ of the relationship shown in the previous panel ($\\Delta V_{TA}$ vs response size contrast across excitatory neurons). Each marker corresponds to one network and target stimulus. Red markers indicate a significant correlation (p < 0.05). Inset histograms show the distribution of the corresponding values, with orange portions indicating the 95\% confidence interval of the median, calculated by bootstrap with 10000 iterations.
\textbf{f} In the sample network, contribution of the contrast in $V_{TA}$ and $V_m$ to bins with increased firing in dev (i.e., positive/red bins in the spike probability contrast, panel a), weighted by that same contrast.
\textbf{g} Contribution of $V_{TA}$ relative to the sum of the absolute values of the weighted contributions in panel f.
\textbf{h} Relative contribution of $V_{TA}$ across networks and stimuli, showing median (solid line) and inter-quartile range (shaded area)

@@

Having established that the increased response to A in dev trials was driven by lower thresholds, we next turned to the cause of this reduction. Firstly, plotting $V_{TA}$ at the start of A trials in Figure \ref{fig5}a, we found that most of the network was indeed less adapted in dev, with only a few neurons in the vicinity of B showing higher thresholds. We hypothesized that this could be due to the different non-target stimulation pattern, since B was stimulated more frequently in dev than msc. Indeed, as shown in Figure \ref{fig5}b, the non-target response activity in the dev sequence -- in other words, the response to B -- was concentrated around the site where B was delivered, while in the msc sequence, activity in response to the various non-target stimuli was dispersed across most of the network. Notably, although the comparison is between 400 B trials in dev, and only 100 B trials (alongside 300 C, D and E trials) in msc, only a small handful of neurons responded more to non-target stimulation in dev, most of these very close to the stimulation site of B.

\begin{figure*}%
    \centering
    \includegraphics{fig5.pdf}
    \caption{\textbf{a} $V_{TA}$ at the start of A trials (mean over trials), with each neuron shown in its spatial location. Left and middle column, raw averages for dev and msc sequences, respectively; right column, contrast between dev and msc averages. A and B stimulus locations for this and following figures are highlighted in the left plot; see Figure \ref{fig1} for the locations of the remaining stimuli used in the msc sequence. \textbf{b} Average response (in spikes per trial) to non-target trials (B in dev, B/C/D/E in msc). \textbf{c} Left and middle: Replacement and adaptation components of the non-target response contrast (\textbf{b}, right column), where replacement is the contrast (B msc - C/D/E msc), and adaptation is the contrast (B dev - B msc). Right: Network-wide mean replacement and adaptation values across networks and target stimuli. The adaptation effect is less than zero (t = -9, p = 2.02e-15), and less than the replacement effect (t = 2.3, p = 0.0119), which is not different from zero (t = -0.69, p = 0.246). \textbf{d} Per-neuron relationship between the contrast (dev - msc) in average number of spikes fired in response to non-target stimulation, and the contrast in $V_{TA}$ at the start of target trials, excluding inhibitory neurons, and associated linear regression (r = 0.99, p = 0). \textbf{e} The same relationship as \textbf{d}, but using the medians across excitatory neurons, and plotting one point per network and target stimulus. The values are highly correlated (r = 0.956, p = 5.98e-65)}
    \label{fig5}
\end{figure*}

\textbf{Fig. 5 a} $V_{TA}$ at the start of A trials (mean over trials), with each neuron shown in its spatial location. Left and middle column, raw averages for dev and msc sequences, respectively; right column, contrast between dev and msc averages. A and B stimulus locations for this and following figures are highlighted in the left plot; see Figure \ref{fig1} for the locations of the remaining stimuli used in the msc sequence. \textbf{b} Average response (in spikes per trial) to non-target trials (B in dev, B/C/D/E in msc). \textbf{c} Left and middle: Replacement and adaptation components of the non-target response contrast (\textbf{b}, right column), where replacement is the contrast (B msc - C/D/E msc), and adaptation is the contrast (B dev - B msc). Right: Network-wide mean replacement and adaptation values across networks and target stimuli. The adaptation effect is less than zero (t = -9, p = 2.02e-15), and less than the replacement effect (t = 2.3, p = 0.0119), which is not different from zero (t = -0.69, p = 0.246). \textbf{d} Per-neuron relationship between the contrast (dev - msc) in average number of spikes fired in response to non-target stimulation, and the contrast in $V_{TA}$ at the start of target trials, excluding inhibitory neurons, and associated linear regression (r = 0.99, p = 0). \textbf{e} The same relationship as \textbf{d}, but using the medians across excitatory neurons, and plotting one point per network and target stimulus. The values are highly correlated (r = 0.956, p = 5.98e-65)

We quantified the correlation between the differences in non-target response magnitude and strength of adaptation before target trials across excitatory neurons, and found an almost perfect relationship (Pearson's r = 0.99, n = 800, see Figure \ref{fig5}d). This effect was robust across networks: Greater reduction in the non-target response was closely correlated with a greater reduction in $V_{TA}$ at the start of A trials.

Finally, we examined the cause of the reduced response in non-target trials. Going from the msc sequence to the dev sequence, two things change: Firstly, stimulations with non-target C, D and E are replaced with B, which would cause different responses even in the absence of adaptation, and secondly, the more frequent presentation of B changes adaptation, reducing the average response. To estimate the effect of stimulus replacement and adaptation, respectively, we contrasted the response to B in the msc sequence with the response to other non-target stimuli in the same sequence (replacement), and with the response to B in dev (adaptation). We found that, in this particular network and stimulus pair, adaptation weakly but consistently decreased the response, whereas stimulus replacement reduced many responses more dramatically while permitting larger responses in many neurons in the general vicinity of the stimulus sites of A and B. Across networks, however, the effect of adaptation was more pronounced and more consistent, removing an average of 71 spikes per trial, whereas stimulus replacement had no effect on average.

To conclude, TA plays two complementary roles in deviance detection, affecting the standard and deviant response separately: On the one hand, it reduces the response to the frequently presented standard stimulus, leading to a larger deviant response merely by contrast with the corresponding standard (SSA). On the other hand, due to the overlap in receptive fields, TA acts as blanket suppression in the many-standards control condition, against which the deviant response, relieved from this suppression and encountering a less fatigued network, stands out.

>>>>

\begin{figure*}%
    \centering
    \includegraphics{fig4.pdf}
    \caption{}
    \label{fig4}
\end{figure*}
\textbf{Fig. 4} Responses and threshold adaptation with TA only.
\textbf{a} Target trial average post-stimulus spike histogram, showing trial time along the horizontal axis, and neurons along the vertical, sorted by the time of the first recorded spike across all trials and conditions. Left and middle column, target trials in dev sequence and msc sequence, respectively; right column, contrast between these two.
\textbf{b} Target trial average of the threshold adaptation voltage $V_{TA}$, with axes, neuron order and columns as in panel a.
\textbf{c} Target trial average of the membrane potential $V_m$, with axes, neuron order and columns as in panel a. The first 10 ms of the neurons under direct stimulation are masked out to avoid saturation of the color mask.
\textbf{d} Relationship across excitatory neurons between the contrast (dev - msc) in $V_{TA}$ at the start of target trials and the contrast in average number of spikes fired in target trials, and associated linear regression.
\textbf{e} Mean contrast (dev - msc) in $V_{TA}$ across networks and stimuli, plotted against the Pearson correlation coefficients $\rho$ of the relationship shown in the previous panel ($\\Delta V_{TA}$ vs response size contrast across excitatory neurons). Each marker corresponds to one network and target stimulus. Red markers indicate a significant correlation (p < 0.05). Inset histograms show the distribution of the corresponding values, with orange portions indicating the 95\% confidence interval of the median, calculated by bootstrap with 10000 iterations.
\textbf{f} In the sample network, contribution of the contrast in $V_{TA}$ and $V_m$ to bins with increased firing in dev (i.e., positive/red bins in the spike probability contrast, panel a), weighted by that same contrast.
\textbf{g} Contribution of $V_{TA}$ relative to the sum of the absolute values of the weighted contributions in panel f.
\textbf{h} Relative contribution of $V_{TA}$ across networks and stimuli, showing median (solid line) and inter-quartile range (shaded area)

@@

Having established that the increased response to A in dev trials was driven by lower thresholds, we next turned to the cause of this reduction. A priori, since threshold adaptation is a direct reflection of activity, and since the response to target trials was larger in dev than in msc, the lower thresholds in dev must have been caused by lower activity in non-target trials, i.e. lower responses to B in dev than to B through E in msc. Yet, with activity modulated by thresholds as shown above, we might expect to observe higher dev responses regardless of stimulus. To resolve this apparent paradox, we will turn to an analysis of the spatial arrangement of activity and adaptation.

First, we mapped the average value of $V_{TA}$ at the start of target trials to the spatial location of the neurons, as shown in Figure \ref{fig5}a. In the msc sequence, $V_{TA}$ was roughly evenly distributed across the network, consistent with the random, distributed nature of stimulation. Conversely, in the dev sequence, neurons near the non-target stimulation site (B) were strongly adapted, whereas the majority of the network was less adapted, consistent with the data shown above.
Next, we turned to the likely cause of this difference, the non-target responses (Figure \ref{fig5}b), mapped into space as the average number of spikes fired in each neuron. As expected, the pattern of activity in non-target trials, constituting 80\% of all trials, closely matched that of the thresholds. Notably, although the comparison is between 400 B trials in dev, and only 100 B trials (alongside 300 C, D and E trials) in msc, only a small handful of neurons responded more to non-target stimulation in dev, most of these very close to the stimulation site of B.

Correlating the non-target response and thresholds excitatory neurons (Figure \ref{fig5}c), we found that the reduced non-target response magnitude in dev (t = -7.61, p = 3.5e-12, median = -0.044; one-sided t-test, n=800) almost perfectly predicted the resulting reduction in threshold adaptation levels (Pearson's $\rho$ = 0.99, p = 0).
This finding held across networks and stimuli, as shown in Figure \ref{fig5}d, with a median correlation coefficient of 0.994 and a consistent reduction in the average non-target response between msc and dev sequences (t = -11.6, p = 1.46e-21, median = -0.058; one-sided t-test, n=120), clearly validating our notion that the non-target trial response is the primary driver of threshold adaptation.

\begin{figure*}%
    \centering
    \includegraphics{fig5.pdf}
    \caption{}
    \label{fig5}
\end{figure*}
\textbf{Fig. 5} Non-target response magnitude determines thresholds.
\textbf{a} Sample network $V_{TA}$ at the start of A trials (mean over trials), with each neuron shown in its spatial location. Left and middle column, raw averages for dev and msc sequences, respectively; right column, contrast between dev and msc averages. A and B stimulus locations for this and following figures are highlighted in the left plot; see Figure \ref{fig1} for the locations of the remaining stimuli used in the msc sequence.
\textbf{b} Average response (in spikes per trial) to non-target trials (B in dev, B, C, D, and E in msc). Columns are arranged as in panel a.
\textbf{c} Relationship across excitatory neurons between the contrast (dev - msc) in average number of spikes fired in response to non-target stimulation and the contrast in $V_{TA}$ at the start of target trials, and associated linear regression.
\textbf{d} Mean contrast (dev - msc) in the non-target response magnitude across networks and stimuli, plotted against the Pearson correlation coefficients $\rho$ of the relationship shown in the previous panel (non-target response size contrast vs $\\Delta V_{TA}$ across excitatory neurons). Each marker corresponds to one network and target stimulus. Red markers indicate a significant correlation (p < 0.05). Inset histograms show the distribution of the corresponding values, with orange portions indicating the 95\% confidence interval of the median, calculated by bootstrap with 10000 iterations

\begin{figure*}%
    \centering
    \includegraphics{fig6.pdf}
    \caption{}
    \label{fig6}
\end{figure*}
\textbf{Fig. 6} Stimulus replacement and adaptation both lead to reduced dev non-target response.
\textbf{a} Non-target response contrasts: The replacement contrast (top right) compares the response to B in msc (top left) to all non-target responses in msc (top center). The adaptation contrast (bottom right) compares the response to B in dev (bottom left) to the response to B in msc (bottom center). Both contrasts together add up to the non-target contrast (dev - msc) shown in Figure \ref{fig5}b. Note that the contrast column (right) shares a color scale to allow a direct comparison.
\textbf{b} Mean replacement (left) and adaptation (right) contrasts across networks and stimuli, as well as the mean replacement contrast filtered for neurons near B, i.e., neurons within one excitatory synapse of the neurons under direct stimulation of B (center).
\textbf{c} Mean $V_{TA}$ contrast (dev - msc) at the start of A trials, filtered for neurons near B. Each marker corresponds to one network; the sample network is highlighted in red. The histogram in the background shows the distribution of the data, with the orange portion indicating the 95\% confidence interval of the median, calculated by bootstrap with 10000 iterations

Finally, we examined the cause of the reduced response in non-target trials. Going from the msc sequence to the dev sequence, two things change: Firstly, stimulations with non-target C, D and E are replaced with B, which would cause different responses even in the absence of adaptation. Secondly, the more frequent presentation of B changes adaptation, likely reducing the average response to B.
In order to separate these two, we split the non-target contrast shown in Figure \ref{fig5}b into two components: a replacement effect, contrasting the response to B in the msc sequence with the response to all non-target stimuli in the same sequence; and an adaptation effect, contrasting the response to B in dev with the response to B in msc. For the sample network, these contrasts are shown in Figure \ref{fig6}a. Unsurprisingly, replacement yielded a decreased response near the stimulation sites for C, D and E, and an increased response in areas near the stimulation site for B. By contrast, adaptation caused a decreased response in virtually the entire network, particularly in the vicinity of the stimulation site for B.

Across networks and stimuli (Figure \ref{fig6}b), we saw a wide range of replacement effects, including both increased and decreased firing, likely as a result of differing sensitivities of the networks to the various non-target stimuli. On average, stimulus replacement did not have a significant effect (t = -0.255, p = 0.799; two-sided t-test). In contrast, adaptation yielded a reduced response on average (t = -9.01, p = 2.02e-15; one-sided t-test), indicating that the more frequent presentation of B indeed lowered the network's response to this stimulus.

Finally, given the clear spatial pattern exhibited by the replacement effect in the sample network, we surmised that adaptation is driven mainly by neurons close to the stimulation site B responding more frequently due to the more frequent presentation of the stimulus. To show this, we first identified all neurons located within a single excitatory synapse of the stimulated neurons (cf. Figure \ref{1}c), and calculated the average replacement effect for these alone. In this subpopulation, replacement yielded a clear response increase, shown in the central box of Figure \ref{fig6}b (t = 8.21, p = 1.5e-13; one-sided t-test). As shown in Figure \ref{fig6}c, this response increase directly translated into elevated thresholds in dev trials compared to msc (t = 5.01, p = 9.59e-07).

Earlier, we showed that the early response is strongly influenced by TA (Figure \ref{fig4}h). Here, we showed that $V_{TA}$ near the stimulation site was increased in dev as a result of the more frequent presentation of B. We posit that this increase in TA affects the early response, reducing the response further downstream and freeing up the rest of the network to respond more strongly to the infrequent target stimulus.

To conclude, TA plays two complementary roles in deviance detection, affecting the standard and deviant response separately: In its first role, as we have shown with the adaptation contrast above, TA reduces the response to frequently presented stimuli regardless of context, leading to a stronger response to infrequently presented stimuli (SSA). Since this happens without the need for large-scale adaptation -- only a small fraction of the network near the stimulation site adapts strongly to frequently presented inputs -- the network as a whole is allowed to relax in the oddball condition. In its second role, TA causes network-wide adaptation in the control condition, where many different stimuli are presented, all of which are infrequently presented and therefore able to evoke large responses. Deviant detection then arises as a relief from this blanket suppression, or as a result of adaptation shifting from the entire network to a small region around the frequently presented standard.
